---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# quickpsy

<!-- badges: start -->
<!-- badges: end -->

quickpsy is an R package developed by [Daniel Linares](http://www.dlinares.org) and [Joan López-Moliner](http://www.ub.edu/viscagroup/joan/) to quickly fit and plot psychometric functions for multiple conditions. 

## Installation

You can install the released version of quickpsy from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("quickpsy")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("danilinares/quickpsy")
```

## Features

- Fits and plots multiple conditions with minimal coding. 
  
- The user does not need to introduce initial parameters for default shapes of the psychometric functions.
  
- Calculates parametric and non-parametric bootstrap confidence intervals. 

- Compares parameters and thresholds for different conditions using bootstrap.
  
- Guess and lapses can be fixed or free as parameters. 
 
- Fits cumulative normal, logistic, weibull functions or any function defined by the user. 
  
- Facilitates the reading of several data files.

### Development version

- Different conditions could share parameters. 

- Easy model comparisons using likelihood ratios or Akaike information criterion. 

## Example

```{r example, fig.height=2, fig.width=6, results='hide'}
library(quickpsy)

fit <- quickpsy(qpdat, phase, resp, grouping = c("participant", "cond"))

plot(fit, color = cond)
```
