---
title: "Basics"
author: "Daniel Linares and Joan López-Moliner"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basics}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
## Reading the files

For the example, download the 3 files in https://github.com/danilinares/quickpsy/tree/master/inst/extdata/example1 and add them to your working directory

The files can be read manually using the R base `read.table` function

```{r, message=FALSE}
library(boot)
library(DEoptim)
library(dplyr)
library(ggplot2)
library(tidyr)
library(quickpsy)
dat1 <- read.table('aa1.txt', head = T)
dat2 <- read.table('bb1.txt', head = T)
dat3 <- read.table('cc2.txt', head = T)

dat1 <- dat1 %>% mutate(subject = 'aa')
dat2 <- dat2 %>% mutate(subject = 'bb')
dat3 <- dat3 %>% mutate(subject = 'cc')

dat <- rbind(dat1, dat2, dat3)
```

or using the *quickpsy* function `quickreadfiles`

```{r}
dat <- quickreadfiles(subject = c('aa', 'bb', 'cc'),
                      session = c('1', '2'))
```

## Fitting 

We use the main function of *quickpsy*, which has the name of the package, to do the fit. We introduce the data, the name of the explanatory variable, the name of the response variable and the name of the grouping variable. 

```{r}
fit <- quickpsy(dat, phase, resp, grouping=.(subject)) 
# we can also introduce the name of the arguments: 
# fit <- quickpsy(d = dat, x = phase, k = resp, grouping=.(subject))
```

`quickpsy` creates an list of class *quickpsy* with all the results. We can use `names` to obtain the names of the output variables

```{r}
names(fit)
```

To obtain the parameters, for example, we can do
```{r}
parameters <- fit$para
parameters
```

or to get the confidence intervals of the parameters
```{r}
fit$paraci
```

If, for example, we want to have the parameters and the confidence intervals in the same data frame we can just `merge`

```{r}
merge(fit$para, fit$paraci)
```

## Plotting

To plot the curves

```{r, fig.align='center', fig.width=5}
plot1 <- plotcurves(fit)
plot1
```

plot1 is a *ggplot* that we can modify accordingly

```{r, fig.align='center', fig.width=5}
plot1 + theme_classic() + ylab('Proportion')
```


