---
title: "guesses_and_lapses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{guesses_and_lapses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `guess` argument of `quickpsy` allows you to set the leftward asymptote of the psychometric function. 

The default value is `0`, but it can be changed:

```{r fig.height=2, fig.width=6}
library(quickpsy)
fit <- quickpsy(qpdat, phase, resp, grouping = c("participant", "cond"), 
                guess = 0.05, bootstrap = "none") 

plot(fit, color = cond)
```

The leftward asymptote can also be free as a parameter (it wouldn't be a guess rate in this case)

```{r fig.height=2, fig.width=6}
fit <- quickpsy(qpdat, phase, resp, grouping = c("participant", "cond"),
                guess = TRUE, B = 10) # only 10 bootstrap samples

plot(fit, color = cond)
```




The `lapses` argument of `quickpsy` works as the `guess` argument. 

We can combined guesses and lapses. For example, for a task with a fixed guess rate of 50% and a variable lapse rate parameter we can do 

```{r fig.height=2, fig.width=3}

library(ggplot2)

n <- 100
x <- c(.2, .4, .6, .8, 1) 
k <- c(59, 56, 69, 92, 94) 
y <- k / n
dat <- data.frame(x, k, y)

fit <- quickpsy(dat, x, k, n, guess = 0.5, lapses = TRUE, bootstrap = "none")

plot(fit) + ylim(0.25, 1)
```

Observe that the lapse rate parameter is now the `p3` parameter (`p1` and `p2` correspond to the position and scale parameters respectively)

```{r}
fit$par
```
If both the leftward and the rightward asymptotes are free, they will appear as the `p3` and `p4` parameters respectively. 

