---
title: "Psychometric functions and lapses"
author: "Daniel Linares and Joan López-Moliner"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Psychometric functions and lapses}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

Here we use *quickpsy* to illustrate the problem of fitting psychometric functions when observers make lapses.

## Lapses can bias threshold estimation, a simple example

Wichmann and Hill (2001) showed that when lapses occur, the estimated threshold could be biased. Consider the following toy example
```{r, warning=FALSE, message=FALSE, fig.align='center', fig.width=5}
library(dplyr)
library(ggplot2)
library(quickpsy)

x <- seq(0,420,60)
k <- c(0, 0, 4, 18, 20, 20, 19, 20)
dat <- data.frame(x, k, n = 20)

fitWithoutLapses <- quickpsy(dat, x, k, n, prob = .75) 

curvesWithoutLapses <- fitWithoutLapses$curves %>% mutate(cond = 'Without Lapses')

pWithout <- ggplot()+
  geom_point(data = fitWithoutLapses$averages, aes(x = x, y = y)) +
  geom_line(data = curvesWithoutLapses,
            aes(x = x, y = y, color = cond)) +
  geom_linerange(data = fitWithoutLapses$thresholds, 
                 aes(x = thre, ymin = 0, ymax = prob), lty =2) 
pWithout
```

Because of the lapses, the fit doesn't seem good and the threshold estimation biased. The problem is not that the likelihood is not correctly maximized. This *weird* fit is the one with the maximum likelihood, given the lapses (Wichmann and Hill, 2001). To be more secure, that the optimization process is not reaching a local maximum, we repeate the fit using Differential Evolution algoriths and obtain the same result

```{r, warning=FALSE, message=FALSE, results='hide', fig.align='center', fig.width=5}
fitWithoutLapses2 <- quickpsy(dat, x, k, n, prob = .75, optimization = 'DE',
                              pini = list(c(1, 500), c(1, 500), c(0,1)),
                              bootstrap = 'none') 

curvesWithoutLapses2 <- fitWithoutLapses2$curves %>% mutate(cond = 'Without Lapses')

pWithout2 <- ggplot()+
  geom_point(data = fitWithoutLapses2$averages, aes(x = x, y = y)) +
  geom_line(data = curvesWithoutLapses,
            aes(x = x, y = y, color = cond)) +
  geom_linerange(data = fitWithoutLapses2$thresholds, 
                 aes(x = thre, ymin = 0, ymax = prob), lty =2) 
pWithout2
```

Wichmann and Hill (2001) reported that allowing the upper assymptote to vary  — that is, fitting the lapses — eliminates the bias for the threshold, which seems to work in our example

```{r, warning=FALSE, fig.align='center', fig.width=5}
fitWithLapses <- quickpsy(dat, x, k, n, prob = .75, lapses = T) 
curvesWithLapses <- fitWithLapses$curves %>% mutate(cond = 'With Lapses')

pWithoutWith <- pWithout +
  geom_line(data = curvesWithLapses,
            aes(x = x, y = y, color = cond)) +
  geom_linerange(data = fitWithLapses$thresholds, 
                 aes(x = thre, ymin = 0, ymax = prob), lty =2) 
pWithoutWith
```

## Lapses can bias threshold estimation, a second example

```{r, warning=FALSE, message=FALSE, fig.align='center', fig.width=5}
data(qpdat)

fitWithoutLapses <- quickpsy(qpdat, phase, resp,grouping = c(ecc, interval ,participant)) 

plotcurves(fitWithoutLapses)
```

Some fits, like the upper left panel for the blue dots, don't seem  good due to lapses, so let's allow the lapses to vary

```{r, warning=FALSE, message=FALSE, error=TRUE}

fitWithLapses <- quickpsy(qpdat, phase, resp, grouping = c(ecc, interval ,participant), lapses  = T) 

```

We obtain an error in the bootstrap. To understand the error we plot the results without bootstraping

```{r, warning=FALSE, message=FALSE, fig.align='center', fig.width=5}
data(qpdat)

fitWithLapses <- quickpsy(qpdat, phase, resp, grouping = c(ecc, interval ,participant), lapses  = T, bootstrap = 'none') 

plotcurves(fitWithLapses) + ylim (0,1) # we include ggplot ylim to be able to visualize what is going on
```

## Lapses bias threshold estimation

Wichmann and Hill (2001) showed that when lapses occur, the estimated threshold is biased. Using simulations like theirs (see the original paper for details), here we replicate this result.

Wichmann and Hill used 7 sampling squemes defined as follow
```{r, message=FALSE, fig.align='center', fig.width=5}


paraweibull <- c(10, 3)
create_xs <- function(i, f) data.frame(squeme = i,  y = f, x = inv_weibull_fun(f, paraweibull))
s <- list()
s[[1]] <- create_xs(1, c(.3, .4, .48, .52, .6, .7))
s[[2]] <- create_xs(2, c(.1, .3, .4, .6, .7, .9))
s[[3]] <- create_xs(3, c(.3, .44, .7, .8, .9, .98))
s[[4]] <- create_xs(4, c(.1, .2, .3, .4, .5, .6))
s[[5]] <- create_xs(5, c(.08, .18, .28, .7, .85, .99))
s[[6]] <- create_xs(6, c(.3, .4, .5, .6, .7, .99))
s[[7]] <- create_xs(7, c(.34, .44, .54, .8, .9, .98))
s <- do.call('rbind', s) 

ggplot(s, aes(x = x, y = squeme, color = factor(squeme))) + 
  geom_point() + geom_line() + 
  labs(color = 'Sampling scheme') + theme(legend.position = 'top')
```

They generated parametric bootstrap samples of data coming from a psychometric function with the shape of a weibull function with parameters $10$ and $3$, $guess=0.5$ and variable $\lambda$ (directly related to the lapses). 

For each value of the independent variable $x$, we simulated 160 trials (in the original paper 20, 40 and 80 trials were also used). 

```{r}
create_sim_dat <- function(d) {
  psychometric_fun <- create_psy_fun(weibull_fun, .5, d$lambda)
  ypred <- psychometric_fun(d$x, paraweibull)
  k <- rbinom(length(d$x), d$n, ypred)
  data.frame(x = d$x, k = k, n = d$n , y = k/d$n)
}

simdat <- merge(s, expand.grid(n = 160, sample = 1:10, lambda = seq(0,.05, .01))) %>%
  group_by(squeme, n, sample, lambda) %>% do(create_sim_dat(.)) 
# we use only 30 samples to increase speed

```

To illustrate the simulated data, we plot the first sample for each sampling scheme (columns) and each value of $\lambda$ (rows)

```{r, fig.align='center',fig.width=7,fig.height=7}
p <- ggplot(simdat %>% filter(sample == 1)) + facet_grid(lambda ~ squeme) + geom_point(aes(x = x, y = y))
p
```

We use quickpsy to fit a weibull function to each condition using $\lambda = 0$

```{r, warning=FALSE}
fit <- quickpsy(simdat, x, k, n, within = .(squeme, lambda, sample), 
                fun = weibull_fun, bootstrap = 'none', guess =.5, lapses = 0) 
# we use 'none' for the bootstap because we are doing an explicit bootstrap  (we are fitting by sample)
# that is, we don't need to bootstrap each of our samples
```

and for illustration purposes show the fitted psychometric function for the first sample 

```{r,fig.align='center',fig.width=7,fig.height=7}
p + geom_line(data = curves(fit, xmin = 0, xmax = 18) %>% filter(sample == 1),
              aes(x = x, y =y)) 
# we use the function curves instead of fit$curves because we want to define the range of the curves manually
```

Replicating Wichmann and Hill, we show that the thresholds are overstimated as $\lambda$ increases (they showed that the slope is also biased)
```{r,fig.align='center',fig.width=4}
thre <- fit$thresholds %>% group_by(squeme, lambda) %>%
  summarise (threshold = mean(thre))

real_threshold <- inv_weibull_fun((.75 - .5) / (1 - .5 - 0), paraweibull)

ggplot(thre) + 
  geom_point(aes(x = lambda, y = threshold, color = factor(squeme)), size = 4) +
  geom_hline(yintercept = real_threshold, lty = 2) 
```

## Fitting lapses does not reduce the bias in general

Wichmann and Hill (2001) reported that allowing $\lambda$ to vary within a given small window eliminates the bias for the threshold, but consistent with Prins (2012) we were not able to replicate this finding 
```{r,fig.align='center',fig.width=4}
fit_lapses <- quickpsy(simdat, x, k, n, within = .(squeme, lambda, sample), 
                fun = weibull_fun, bootstrap = 'none', guess =.5, lapses = T,
                pini = list(c(1,30), c(1,10), c(0,.05))) # we introduce the min and max value for each parameter

thre_lapses <- fit_lapses$thresholds %>% group_by(squeme, lambda) %>%
  summarise (threshold = mean(thre))

ggplot(thre_lapses) + 
  geom_point(aes(x = lambda, y = threshold, color = factor(squeme)), size = 4) +
  geom_hline(yintercept = real_threshold, lty = 2) 
```

## But for some data, fitting lapses does eliminate the bias 

We illustrate this, with an extreme example

```{r, warning=FALSE, fig.align='center', fig.width=5}
x <- seq(0,420,60)
k <- c(0, 0, 4, 18, 20, 20, 19, 20)
dat <- data.frame(x, k, n = 20)

fitWithoutLapses <- quickpsy(dat, x, k, n) 
fitWithLapses <- quickpsy(dat, x, k, n, lapses = T) 

curvesWithoutLapses <- fitWithoutLapses$curves %>% mutate(cond = 'Without Lapses')
curvesWithLapses <- fitWithLapses$curves %>% mutate(cond = 'With Lapses')

ggplot()+
  geom_point(data = fitWithoutLapses$averages, aes(x = x, y = y)) +
  geom_line(data = curvesWithoutLapses,
            aes(x = x, y = y, color = cond)) +
  geom_line(data = curvesWithLapses,
            aes(x = x, y = y, color = cond)) 
```

When the number of parameters to be fitted is 3 or more the minization sometimes reaches local minima. To try to avoid that, differential evolution algorithms could be used (DEoptim)

```{r, results='hide', fig.align='center', fig.width=5}
fitWithLapses <- quickpsy(dat, x, k, n, lapses = T, optimization = 'DE', bootstrap = 'none',
                          pini = list(c(10, 1000), c(1,100), c(0, .1))) 

curvesWithLapses <- fitWithLapses$curves %>% mutate(cond = 'With Lapses')

ggplot()+
  geom_point(data = fitWithoutLapses$averages, aes(x = x, y = y)) +
  geom_line(data = curvesWithoutLapses,
            aes(x = x, y = y, color = cond)) +
  geom_line(data = curvesWithLapses,
            aes(x = x, y = y, color = cond)) 

```

## References

Prins, N. (2012). The psychometric function: The lapse rate revisited. Journal of Vision, 12(6), 25–25. 

Wichmann, F. A., & Hill, N. J. (2001). The psychometric function: I. Fitting, sampling, and goodness of fit. Perception and Psychophysics, 63(8), 1293–1313.


