---
title: "model_selection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{model_selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


To perform model selection, the dev version of quickspy is necessary. 

Model selection is quite flexible in `quickpsy` as in the `fun` argument you can introduce any arbitrary set of functions with all possible combinations of shared parameters (see the last subsection of the functions section) 

To perform model selection, `quickpsy` uses likelihood ratio tests and the Akaike Information Criterion (see References). 

## Model 1

Let's fit a model for each participant in which the slope does not change across conditions 

```{r fig.height=2, fig.width=6, message=FALSE}
library(quickpsy)
library(dplyr)

# the shared parameter is p[2]
fun_1_df <- tibble(cond = c("cond1", "cond2"), 
                   fun = c(function(x, p) pnorm(x, p[1], p[2]), 
                           function(x, p) pnorm(x, p[3], p[2])))

fit_1 <- quickpsy(qpdat, phase, resp, 
                  grouping = c("participant", "cond"), 
                  fun = fun_1_df, 
                  parini = c(-100, 50, -100), 
                  bootstrap = "none") 

plot(fit_1, color = cond)
```

We can look at the likelihoods of the models 

```{r}
fit_1$logliks
```

or the AICs

```{r}
fit_1$aic
```

## Model 2

Now let's fit a model in which for each participant the slope can change across conditions

```{r fig.height=2, fig.width=6, message=FALSE}
fun_2_df <- tibble(cond = c("cond1", "cond2"), 
                   fun = c(function(x, p) pnorm(x, p[1], p[2]), 
                           function(x, p) pnorm(x, p[3], p[4])))

fit_2 <- quickpsy(qpdat, phase, resp, 
                  grouping = c("participant", "cond"), 
                  fun = fun_2_df, 
                  parini = c(-100, 50, -100, 50), 
                  bootstrap = "none") 

plot(fit_2, color = cond)
```

## Model selection: likelihood ratio test

The `quickpsy` function `model_selection_lrt` performs the likelihood ratio tests using the chi square distribution. The inputs are the likelihoods of each model. 

```{r}
model_selection_lrt(fit_1$logliks, fit_2$logliks)
```

Using the default 5% criterion of significance, the Model 2 (allowing a different slope for each condition) does not improve the fit for any participant. There is some evidence, however, that for Participant 1 the Model 2 might be better. 

## Model selection: Akaike Information Criterion

To perform the model selection using `quickpsy`, you need to introduce in the  `model_selection_aic` function the AICs of each model.

```{r}
model_selection_aic(fit_1$aic, fit_2$aic)
```
Using Akaike Information Criterion the Model 1 is preferred for Participant 2 and 3 and the Model 2 is preferred for Participant 1. The `p` indicates the relative probability of each model. 


## References

Kingdom FAA, Prins N. 2016. Psychophysics: A Practical Introduction. Elsevier Science.

Knoblauch, K., & Maloney, L. T. (2012). Modeling psychophysical data in R (Vol. 32). Springer Science & Business Media.

Prins, N. (2018). Applying the model-comparison approach to test specific research hypotheses in psychophysical research using the Palamedes toolbox. Frontiers in psychology, 9, 1250.



